---
title: "R Notebook"
output: html_notebook
---

Ryan E. Kelly
CS 668 - Analytics Capstone
Prof Krystyn Gutu
4 December 2024
#FINAL CODE REVIEW




# DATASET 1 -  New York State GDP Data
Source: Federal Reserve Bank (St. Louis)
https://fred.stlouisfed.org/series/NYNQGSP
```{r}

library(fredr)

fredr_set_key("0200e442540c7e00b0062bbabaf750dd")

ny_gdp_data <- fredr(series_id = "NYNGSP")
colnames(ny_gdp_data) <- c("date", "series_id", "gdp_value", "realtime_start", "realtime_end")

ny_gdp_data$year <- as.numeric(format(ny_gdp_data$date, "%Y"))
ny_gdp_data$gdp_value <- ny_gdp_data$gdp_value / 1e6  # Converts to billions
ny_gdp_data <- ny_gdp_data[, c("year", "gdp_value")]

str(ny_gdp_data)
summary(ny_gdp_data)

```


```{r}
windowsFonts(A = windowsFont("Times New Roman"))
par(family = "A", mar = c(5, 5, 4, 2) + 0.1)

year_range <- paste(min(ny_gdp_data$year), max(ny_gdp_data$year), sep = " - ")

plot(
  ny_gdp_data$year, ny_gdp_data$gdp_value,
  type = "b",
  col = "black",
  xlab = "Calendar Year",
  ylab = "GDP Value (Billions $)",
  main = paste("New York State Nominal GDP CY (", year_range, ")", sep = ""),
  pch = 19,
  lwd = 1,
  las = 1,
  cex.axis = 0.8,
  cex.lab = 1,
  xaxt = "n",  # Turn off default x-axis
  yaxt = "n"   # Turn off default y-axis to customize
)

year_ticks <- seq(min(ny_gdp_data$year), max(ny_gdp_data$year), by = 1)

# Add the custom x-axis with rotated labels
axis(
  side = 1,
  at = year_ticks,
  labels = year_ticks,
  las = 2,     
  cex.axis = 0.8
)

# Add y-axis with "B" appended
y_ticks <- axTicks(2)  # Get default positions for y-axis ticks
axis(2, at = y_ticks, labels = paste0(y_ticks, "B"), las = 1)

# Add horizontal grid lines (ny = 10 for example)
grid(nx = NA, ny = 10, col = "gray", lty = "dotted")

# Add vertical lines for every year on the x-axis
for(x_val in year_ticks) {
  abline(v = x_val, col = "gray", lty = "dotted")
}

# Add data source at the bottom of the plot
mtext("Data Source: Federal Reserve Bank of St. Louis", side = 1, line = 3, cex = 0.8, adj = -0.25)


```

# DATASET 2 -  New York State Unemployment Data
Source: Federal Reserve Bank (St. Louis)
https://fred.stlouisfed.org/series/NYURN

```{r}
library(fredr)

fredr_set_key("0200e442540c7e00b0062bbabaf750dd")
ny_unemployment_data_raw <- fredr(series_id = "NYUR")



ny_unemployment_df <- data.frame(
  year = as.numeric(format(ny_unemployment_data_raw$date, "%Y")),
  period = format(ny_unemployment_data_raw$date, "%m"),  # month period
  period_name = format(ny_unemployment_data_raw$date, "%B"),
  unemployment_rate = ny_unemployment_data_raw$value
)

ny_unemployment_df <- subset(ny_unemployment_df, period != "13")

```

```{r}
# Creation of fiscal year variable based on NY fiscal year (April 1 - March 31)

ny_unemployment_df$fiscal_year <- ifelse(ny_unemployment_df$period %in% c("01", "02", "03"),
                                         ny_unemployment_df$year - 1,
                                         ny_unemployment_df$year)

# Calendar year averages
ny_annual_unemployment <- aggregate(unemployment_rate ~ year, data = ny_unemployment_df, FUN = mean)
colnames(ny_annual_unemployment)[2] <- "unemployment_CY"

# Fiscal year averages
ny_fiscal_unemployment <- aggregate(unemployment_rate ~ fiscal_year, data = ny_unemployment_df, FUN = mean)
colnames(ny_fiscal_unemployment)[2] <- "unemployment_FY"

ny_unemployment_data <- merge(ny_annual_unemployment, ny_fiscal_unemployment, 
                              by.x = "year", by.y = "fiscal_year", all = TRUE)
```


```{r}
windowsFonts(A = windowsFont("Times New Roman"))
par(family = "A", mar = c(5, 5, 4, 2) + 0.1)

year_range <- paste(min(ny_unemployment_data$year), max(ny_unemployment_data$year), sep = " - ")

plot(
  ny_unemployment_data$year,
  ny_unemployment_data$unemployment_FY,
  type = "o",
  col = "black",
  lwd = 1,
  pch = 19,
  xaxt = "n",
  yaxt = "n",
  ylab = "Average Unemployment Rate (%)",
  xlab = "Fiscal Year",
  main = "New York State Average Unemployment Rate FY (1995-2024)",
  xlim = c(1995, max(ny_unemployment_data$year))   # Start from 1995
)

filtered_years <- ny_unemployment_data$year[ny_unemployment_data$year >= 1995]
axis(1, at = filtered_years, labels = filtered_years, las = 2)
axis(2, las = 1)
grid(nx = length(filtered_years), ny = NULL, col = "lightgray", lty = "dotted")
mtext("Data Source: Federal Reserve Bank of St. Louis", side = 1, line = 3, cex = 0.8, adj = -0.25)

```



DATASET #3 - CPI Data - North Eastern US
Source: US Department of Labor - Bureau of Labor Statistics
https://www.bls.gov/regions/mid-atlantic/news-release/consumerpriceindex_northeast.htm

https://data.bls.gov/timeseries/CUUR0100SA0?amp%253bdata_tool=XGtable&output_view=data&include_graphs=true


```{r}
library(httr)
bls_api_key <- "85f2b89f6fd64218a8ef61270b0e8dc1" 
cpi_series_id <- "CUUR0100SA0" # Northeast US CPI


fetch_cpi_data <- function(start_year, end_year, series_id, api_key) {
  url <- paste0(
    "https://api.bls.gov/publicAPI/v2/timeseries/data/", series_id,
    "?registrationkey=", api_key,
    "&startyear=", start_year, "&endyear=", end_year
  )
  
  response <- GET(url)
  
  if (status_code(response) == 200) {
    result <- content(response, as = "parsed", type = "application/json")$Results$series[[1]]$data
    if (is.null(result)) stop("No data returned from API.")
    return(result)
  } else {
    stop("API request failed with status code: ", status_code(response))
  }
}

# The requests had to be separated to accommodate for size limitations. 
data_1990_2009 <- fetch_cpi_data(1990, 2009, cpi_series_id, bls_api_key)
data_2010_2023 <- fetch_cpi_data(2010, 2023, cpi_series_id, bls_api_key)


# Extract and clean up each dataset with checks for NULLs and empty fields
extract_values <- function(data) {
  years <- sapply(data, function(x) if (!is.null(x$year)) x$year else NA)
  periods <- sapply(data, function(x) if (!is.null(x$period)) x$period else NA)
  period_names <- sapply(data, function(x) if (!is.null(x$periodName)) x$periodName else NA)
  cpi_values <- sapply(data, function(x) if (!is.null(x$value)) x$value else NA)
  
  data.frame(
    year = as.numeric(years),
    period = periods,
    period_name = period_names,
    cpi_index = as.numeric(cpi_values)
  )
}

# Apply extraction function
northeast_cpi_1990_2009 <- extract_values(data_1990_2009)
northeast_cpi_2010_2023 <- extract_values(data_2010_2023)

# Combination of data frames
northeast_cpi_df <- rbind(northeast_cpi_1990_2009, northeast_cpi_2010_2023)
head(northeast_cpi_df)
tail(northeast_cpi_df)



```




```{r}

# Aggregate by calendar year
calendar_year_cpi <- aggregate(cpi_index ~ year, data = northeast_cpi_df, FUN = mean)

# Filter data for NYS fiscal year calculation and check if non-empty
fiscal_year_data <- subset(northeast_cpi_df, period_name %in% c("April", "May", "June", "July", "August", "September", "October", "November", "December", "January", "February", "March"))
if (nrow(fiscal_year_data) > 0) {
  # Aggregate for NYS Fiscal Year (April 1 - March 31)
  fiscal_year_cpi <- aggregate(cpi_index ~ year, data = fiscal_year_data, FUN = function(x) mean(x, na.rm = TRUE))
  
  # Adjust fiscal year labels to start in the previous year
  fiscal_year_cpi$year <- fiscal_year_cpi$year - 1
  
  # Merge Calendar Year and Fiscal Year data into a single dataframe
  ne_cpi_data <- merge(calendar_year_cpi, fiscal_year_cpi, by = "year", all = TRUE)
  colnames(ne_cpi_data) <- c("year", "cpi_calendar_year", "cpi_fiscal_year")
  
  head(ne_cpi_data)
  summary(ne_cpi_data)
} else {
  warning("Fiscal year data is empty after subsetting. Check period_name values.")
}


```



```{r}
par(las = 1)
windowsFonts(A = windowsFont("Times New Roman"))
par(family = "A", mar = c(5, 5, 4, 2) + 0.1)

cpi_range <- range(ne_cpi_data$cpi_calendar_year, na.rm = TRUE)
ylim_upper <- cpi_range[2] + 10

# Determine the filtered years starting from 1995
year_ticks <- sort(unique(ne_cpi_data$year))
year_ticks <- year_ticks[year_ticks >= 1995]

plot(
  ne_cpi_data$year, ne_cpi_data$cpi_calendar_year,
  type = "o",
  col = "black", pch = 19, lty = 1, lwd = 1,
  xlab = "Calendar Year", ylab = "CPI Index",
  main = "Northeast Consumer Price Index (CPI) - CY(1995-2023)",
  ylim = c(cpi_range[1], ylim_upper),
  xlim = c(1995, max(year_ticks)),
  xaxt = "n"
)

# Add rotated x-axis labels for years >= 1995
axis(1, at = year_ticks, labels = year_ticks, las = 2, cex.axis = 0.8)

# Add horizontal grid lines
grid(nx = NA, ny = NULL, col = "gray", lty = "dotted")

# Add vertical lines for each year >= 1995
for (x_val in year_ticks) {
  abline(v = x_val, col = "gray", lty = "dotted")
}

# Add data source at the bottom-left
mtext("Data Source: U.S. Department of Labor - Bureau of Labor Statistics", 
      side = 1, line = 4, cex = 0.8, adj = -0.40)



```
```{r}

windowsFonts(A = windowsFont("Times New Roman"))
par(family = "A", mar = c(5, 5, 4, 2) + 0.1)

cpi_range_fy <- range(ne_cpi_data$cpi_fiscal_year, na.rm = TRUE)
ylim_upper_fy <- cpi_range_fy[2] + 10

year_ticks <- sort(unique(ne_cpi_data$year))
year_ticks <- year_ticks[year_ticks >= 1995]

plot(
  ne_cpi_data$year, ne_cpi_data$cpi_fiscal_year,
  type = "o",
  col = "black", pch = 19, lty = 1, lwd = 1,
  xlab = "Fiscal Year", ylab = "CPI Index",
  main = "Northeast Consumer Price Index (CPI) - FY(1995-2022)",
  ylim = c(cpi_range_fy[1], ylim_upper_fy),
  xlim = c(1995, max(year_ticks)),
  xaxt = "n"
)


axis(1, at = year_ticks, labels = year_ticks, las = 2, cex.axis = 0.8)

grid(nx = NA, ny = NULL, col = "gray", lty = "dotted")

for (x_val in year_ticks) {
  abline(v = x_val, col = "gray", lty = "dotted")
}

mtext("Data Source: U.S. Department of Labor - Bureau of Labor Statistics",
      side = 1, line = 4, cex = 0.8, adj = -0.40)


```

# DATASET 4 - New York State Tax Projections
Source: New York State Department of Finance
API Call
https://www.tax.ny.gov/research/stats/statistics/stat_fy_collections.htm


```{r}
library(httr)
library(jsonlite)
library(dplyr)

base_url <- "https://data.ny.gov/resource/cpjr-ezcj.json"

# This offset is necessary because the NYS API for tax pulls has a limit of 1000 lines. 
# The tax data has approximately 1,600 lines. This method is recommended by the NYS data contractor, and it doesn't require an API key!

fetch_data <- function(offset, limit = 1000) {
  url <- paste0(base_url, "?$limit=", limit, "&$offset=", offset)
  response <- GET(url)
  
  if (http_type(response) != "application/json") {
    stop("Failed to fetch data: ", status_code(response))
  }
  
  content <- content(response, as = "text", encoding = "UTF-8")
  data <- fromJSON(content, flatten = TRUE)  # flatten nested data for simplicity
  
  if (is.null(data) || length(data) == 0) {
    return(NULL)  # Return NULL if data is empty
  }
  
  return(as.data.frame(data))
}

# Initialize variables
ny_tax_data_raw <- data.frame()  # Renamed to ny_tax_data_raw
offset <- 0
limit <- 1000

# Loop to fetch all data
repeat {
  data <- fetch_data(offset, limit)
  
  # Break if no data was returned
  if (is.null(data) || nrow(data) == 0) break
  
  # Bind the new data to the existing data frame
  ny_tax_data_raw <- bind_rows(ny_tax_data_raw, data)
  
  # Update offset for next iteration
  offset <- offset + limit
}

# Check the final data frame
head(ny_tax_data_raw)  # Show the first few rows
summary(ny_tax_data_raw)


```


```{r}
# Ensure that amount_collected is numeric
ny_tax_data_raw$amount_collected <- as.numeric(ny_tax_data_raw$amount_collected)

# Aggregate amount_collected by fiscal_year for each specified tax category
sales_excise_use <- aggregate(amount_collected ~ fiscal_year, 
                              data = ny_tax_data_raw[ny_tax_data_raw$tax_category == "Sales, Excise and Use", ], 
                              sum, na.rm = TRUE)
other_taxes_fees <- aggregate(amount_collected ~ fiscal_year, 
                              data = ny_tax_data_raw[ny_tax_data_raw$tax_category == "Other Taxes and Fees", ], 
                              sum, na.rm = TRUE)
corporation_business <- aggregate(amount_collected ~ fiscal_year, 
                                  data = ny_tax_data_raw[ny_tax_data_raw$tax_category == "Corporation and Business", ], 
                                  sum, na.rm = TRUE)
personal_income <- aggregate(amount_collected ~ fiscal_year, 
                             data = ny_tax_data_raw[ny_tax_data_raw$tax_category == "Personal Income", ], 
                             sum, na.rm = TRUE)
property_transfer <- aggregate(amount_collected ~ fiscal_year, 
                               data = ny_tax_data_raw[ny_tax_data_raw$tax_category == "Property Transfer", ], 
                               sum, na.rm = TRUE)
total_collected <- aggregate(amount_collected ~ fiscal_year, 
                             data = ny_tax_data_raw[ny_tax_data_raw$tax_category == "Total", ], 
                             sum, na.rm = TRUE)

# Merge all totals by fiscal_year
ny_tax_data <- Reduce(function(x, y) merge(x, y, by = "fiscal_year", all = TRUE), 
                      list(sales_excise_use, other_taxes_fees, 
                           corporation_business, personal_income, 
                           property_transfer, total_collected))


colnames(ny_tax_data) <- c("fiscal_year", "sales_excise_use", 
                           "other_taxes_fees", "corporation_business", 
                           "personal_income", "property_transfer", 
                           "total_collected")

print(ny_tax_data)


```

```{r}
# Check for missing totals and calculate them for fiscal years 1995 to 2002
for (i in 1995:2002) {
  # Check if total_collected is missing or NA for the given fiscal year
  if (is.na(ny_tax_data$total_collected[ny_tax_data$fiscal_year == i])) {
    # Sum across the other categories for this fiscal year
    ny_tax_data$total_collected[ny_tax_data$fiscal_year == i] <- 
      with(ny_tax_data[ny_tax_data$fiscal_year == i, ],
           sales_excise_use + other_taxes_fees + 
           corporation_business + personal_income + 
           property_transfer)
  }
}

print(ny_tax_data)


```



```{r}
windowsFonts(A = windowsFont("Times New Roman"))
par(family = "A", mar = c(5, 5, 4, 2) + 0.1, las = 1)

barplot_values <- barplot(
  ny_tax_data$total_collected / 1e9,
  names.arg = ny_tax_data$fiscal_year,
  col = "lightblue",
  xlab = "Fiscal Year",
  ylab = "Total State Collections (Billions $)",
  main = "New York State Total Tax Revenues by FY (1995-2024)",
  las = 2,
  ylim = c(0, max(ny_tax_data$total_collected / 1e9) * 1.1),
  cex.names = 0.8,
  yaxt = "n"  # Turn off default y-axis to customize
)

# Add a custom y-axis with "B" appended to values
y_ticks <- axTicks(2)  # Get default tick positions for y-axis
axis(2, at = y_ticks, labels = paste0(y_ticks, "B"), las = 1)

grid(nx = NA, ny = NULL, col = "gray")

abline(
  h = seq(20, max(ny_tax_data$total_collected / 1e9), by = 20),
  col = "gray", lty = "dotted"
)

box()

# Add data source at the bottom-left
mtext("Data Source: NYS Department of Finance", side = 1, line = 3, cex = 0.8, adj = -0.25)




```

##################### STEP 2 - MODEL BUILDING ##########################



```{r}
ny_gdp_data
ny_unemployment_data
ne_cpi_data
ny_tax_data

```

```{r}
# Adding fiscal year column for ease of merging
ny_gdp_data$fiscal_year <- ny_gdp_data$year
ny_unemployment_data$fiscal_year <- ny_unemployment_data$year
ne_cpi_data$fiscal_year <- ne_cpi_data$year

# Merging on fiscal year
ny_tax_forecast_data <- merge(ny_tax_data, ny_gdp_data[, c("fiscal_year", "gdp_value")], by = "fiscal_year", all.x = TRUE)
ny_tax_forecast_data <- merge(ny_tax_forecast_data, ny_unemployment_data[, c("fiscal_year", "unemployment_FY")], by = "fiscal_year", all.x = TRUE)
ny_tax_forecast_data <- merge(ny_tax_forecast_data, ne_cpi_data[, c("fiscal_year", "cpi_fiscal_year")], by = "fiscal_year", all.x = TRUE)


colnames(ny_tax_forecast_data) <- c("fiscal_year", "sales_excise_use", "other_taxes_fees", "corporation_business", 
                                    "personal_income", "property_transfer", "total_collected", "gdp_value", 
                                    "unemployment", "cpi")

# Impute missing values for historical data - these values were back filled from 1997 - I could have also manually put these in, but for ease of access with the API and the need to keep this consistant, I am using this approach. This should not greatly impact the model in any way. I will revist this next code review. 
ny_tax_forecast_data$gdp_value[ny_tax_forecast_data$fiscal_year == 1995] <- ny_tax_forecast_data$gdp_value[ny_tax_forecast_data$fiscal_year == 1997]
ny_tax_forecast_data$gdp_value[ny_tax_forecast_data$fiscal_year == 1996] <- ny_tax_forecast_data$gdp_value[ny_tax_forecast_data$fiscal_year == 1997]
ny_tax_forecast_data$cpi[ny_tax_forecast_data$fiscal_year == 2023] <- ny_tax_forecast_data$cpi[ny_tax_forecast_data$fiscal_year == 2022]

# Prepare the 2024 test data and handle missing values
test_data_2024 <- subset(ny_tax_forecast_data, fiscal_year == 2024)

# Forward-fill missing values with 2023 values for 2024 predictions. (Only 1 value is being forward filled for the reason described above. I will revist this next code review)
test_data_2024$gdp_value[is.na(test_data_2024$gdp_value)] <- ny_tax_forecast_data$gdp_value[ny_tax_forecast_data$fiscal_year == 2023]
test_data_2024$cpi[is.na(test_data_2024$cpi)] <- ny_tax_forecast_data$cpi[ny_tax_forecast_data$fiscal_year == 2023]

# View the prepared data
print(head(ny_tax_forecast_data))
print(test_data_2024)


```

##################### MODEL BUILDING ##########################

##################### MODEL ONE - LINEAR REGRESSION ##################### 


```{r}
# Linear Regression Model
training_data <- subset(ny_tax_forecast_data, fiscal_year < 2024)
model <- lm(total_collected ~ gdp_value + unemployment + cpi, data = training_data)

# Predict for 2024
predicted_2024 <- predict(model, newdata = test_data_2024)

# Calculate Evaluation Metrics for Training Data
training_predictions <- predict(model, newdata = training_data)
mae <- mean(abs(training_predictions - training_data$total_collected))
mse <- mean((training_predictions - training_data$total_collected)^2)
r_squared <- summary(model)$r.squared

# Percent Error for 2024 Prediction
actual_2024 <- test_data_2024$total_collected
percent_off <- abs(predicted_2024 - actual_2024) / actual_2024 * 100

# Print Results
cat("Model One: Linear Regression\n")
cat("Training Metrics:\n")
cat("MAE:", round(mae, 2), "\n")
cat("MSE:", round(mse, 2), "\n")
cat("R-squared:", round(r_squared, 2), "\n")
cat("\n2024 Prediction:\n")
cat("Predicted 2024 Tax Revenue:", round(predicted_2024, 2), "\n")
cat("Actual 2024 Tax Revenue:", actual_2024, "\n")
cat("Percent Off:", round(percent_off, 2), "%\n")

```


##################### MODEL ONE - LINEAR REGRESSION - OPTIMZIATION ATTEMPTS  - ORIGINAL REINFORCED #####################

######  ATTEMPT ONE - VIF ######
```{r}
# Linear Regression - VIF Alternative Model

# Check multicollinearity using VIF
library(car)
vif_model <- vif(model)
cat("VIF for each variable in the original model:\n")
print(vif_model)

# Alternative: Remove 'cpi' due to high VIF and refit the model
alternative_model <- lm(total_collected ~ gdp_value + unemployment, data = training_data)

# Evaluate the alternative model
alternative_training_predictions <- predict(alternative_model, newdata = training_data)
alternative_mae <- mean(abs(alternative_training_predictions - training_data$total_collected))
alternative_mse <- mean((alternative_training_predictions - training_data$total_collected)^2)
alternative_r_squared <- summary(alternative_model)$r.squared

# Check VIF again for the alternative model
vif_alternative_model <- vif(alternative_model)
cat("\nVIF for each variable in the alternative model:\n")
print(vif_alternative_model)

# Predict and Calculate Percent Error for 2024
alternative_predicted_2024 <- predict(alternative_model, newdata = test_data_2024)
alternative_percent_off <- abs(alternative_predicted_2024 - actual_2024) / actual_2024 * 100

cat("Alternative Linear Regression\n")
cat("Training Metrics:\n")
cat("MAE:", round(alternative_mae, 2), "\n")
cat("MSE:", round(alternative_mse, 2), "\n")
cat("R-squared:", round(alternative_r_squared, 2), "\n")
cat("\n2024 Prediction:\n")
cat("Predicted 2024 Tax Revenue:", round(alternative_predicted_2024, 2), "\n")
cat("Actual 2024 Tax Revenue:", actual_2024, "\n")
cat("Percent Off:", round(alternative_percent_off, 2), "%\n")

```
######  ATTEMPT TWO - Caret Ridge ######
```{r}
# Linear Regression - Caret Optimization (Ridge Regression)

library(caret)  # For automated optimization within R. This library is updated by CRAN. 

# Define train control and tuning grid
train_control <- trainControl(method = "cv", number = 10)  # 10-folds for cross-validation
ridge_grid <- expand.grid(alpha = 0, lambda = seq(0.001, 1, by = 0.01))  # Alpha = 0 for ridge regression

# Train ridge regression using caret
ridge_model <- train(
  total_collected ~ gdp_value + unemployment + cpi,
  data = training_data,
  method = "glmnet",
  trControl = train_control,
  tuneGrid = ridge_grid
)

# Best hyperparameters
cat("\nCaret Optimization: Ridge Regression\n")
cat("Best Parameters:\n")
print(ridge_model$bestTune)

# Evaluate Ridge Regression on training data
ridge_training_predictions <- predict(ridge_model, newdata = training_data)
ridge_mae <- mean(abs(ridge_training_predictions - training_data$total_collected))
ridge_mse <- mean((ridge_training_predictions - training_data$total_collected)^2)

# Predict for 2024 and calculate percent error
ridge_predicted_2024 <- predict(ridge_model, newdata = test_data_2024)
ridge_percent_off <- abs(ridge_predicted_2024 - actual_2024) / actual_2024 * 100


cat("\nCaret Optimized: Ridge Regression\n")
cat("Training Metrics:\n")
cat("MAE:", round(ridge_mae, 2), "\n")
cat("MSE:", round(ridge_mse, 2), "\n")
cat("\n2024 Prediction:\n")
cat("Predicted 2024 Tax Revenue:", round(ridge_predicted_2024, 2), "\n")
cat("Actual 2024 Tax Revenue:", actual_2024, "\n")
cat("Percent Off:", round(ridge_percent_off, 2), "%\n")

```

######  ATTEMPT THREE - Caret Standard ######
```{r}
library(caret)

# Define train control (cross-validation)
train_control <- trainControl(method = "cv", number = 5)

linear_model_caret <- train(
  total_collected ~ gdp_value + unemployment + cpi,
  data = training_data,
  method = "lm",
  trControl = train_control
)

# Summarize results
print(linear_model_caret)

# Predict for 2024
predicted_2024_caret <- predict(linear_model_caret, newdata = test_data_2024)
percent_off_caret <- abs(predicted_2024_caret - actual_2024) / actual_2024 * 100
cat("Percent Off:", round(percent_off_caret, 2), "%\n")

```

```{r}
# Focus on the last 3 years and the 2024 prediction
focus_years <- c(2021, 2022, 2023, 2024)
focus_actual <- ny_tax_data$total_collected[ny_tax_data$fiscal_year %in% focus_years]

# Extract predictions for all Linear Regression models
focus_linear <- c(focus_actual[1:3], predicted_2024)                     # Original Linear Regression
focus_linear_alternative <- c(focus_actual[1:3], alternative_predicted_2024) # Alternative (VIF-based model)
focus_linear_ridge <- c(focus_actual[1:3], ridge_predicted_2024)         # Ridge Regression
focus_linear_caret <- c(focus_actual[1:3], predicted_2024_caret)         # Caret Optimized


windowsFonts(A = windowsFont("Times New Roman"))
par(family = "A")

plot(focus_years, focus_actual / 1e9, type = "l", col = "black", lwd = 2,
     ylim = c(min(focus_actual) * 0.95 / 1e9, max(focus_actual) * 1.05 / 1e9), 
     xlab = "Fiscal Year", ylab = "Tax Revenue (Billions $)", 
     main = "Linear Regression: Actual vs. All Models", xaxt = "n", yaxt = "n")
grid()


lines(focus_years, focus_linear / 1e9, col = "red", lwd = 2, lty = 2)                    # Original Model
lines(focus_years, focus_linear_alternative / 1e9, col = "blue", lwd = 2, lty = 3)       # Alternative Model (VIF)
lines(focus_years, focus_linear_ridge / 1e9, col = "green", lwd = 2, lty = 4)            # Ridge Regression
lines(focus_years, focus_linear_caret / 1e9, col = "purple", lwd = 2, lty = 5)           # Caret Optimized

axis(1, at = focus_years, labels = focus_years)
axis(2, at = seq(floor(min(focus_actual) / 1e9 * 0.95), 
                 ceiling(max(focus_actual) / 1e9 * 1.05), by = 5), 
     labels = paste0(seq(floor(min(focus_actual) / 1e9 * 0.95), 
                         ceiling(max(focus_actual) / 1e9 * 1.05), by = 5), "B"), 
     las = 1)

legend("topright", legend = c("Actual", "Original Model", "Alternative (VIF)", 
                              "Ridge Regression", "Caret Optimized"), 
       col = c("black", "red", "blue", "green", "purple"), 
       lty = c(1, 2, 3, 4, 5), lwd = 2, cex = 0.7)

```



##################### MODEL TWO - RANDOM FOREST #####################

```{r}

library(randomForest)

# Random Forest Model
training_data_rf <- subset(ny_tax_forecast_data, fiscal_year < 2024)
set.seed(123)
rf_model <- randomForest(total_collected ~ gdp_value + unemployment + cpi, 
                         data = training_data_rf, 
                         ntree = 500, 
                         mtry = 2, 
                         importance = TRUE)

# Predict for 2024
predicted_2024_rf <- predict(rf_model, newdata = test_data_2024)

# Calculate Evaluation Metrics for Training Data
training_predictions_rf <- predict(rf_model, newdata = training_data_rf)
mae_rf <- mean(abs(training_predictions_rf - training_data_rf$total_collected))
mse_rf <- mean((training_predictions_rf - training_data_rf$total_collected)^2)

# Percent Error for 2024 Prediction
actual_2024 <- test_data_2024$total_collected
percent_off_rf <- abs(predicted_2024_rf - actual_2024) / actual_2024 * 100


cat("Model Two: Random Forest Regression\n")
cat("Training Metrics:\n")
cat("MAE:", round(mae_rf, 2), "\n")
cat("MSE:", round(mse_rf, 2), "\n")
cat("\n2024 Prediction:\n")
cat("Predicted 2024 Tax Revenue:", round(predicted_2024_rf, 2), "\n")
cat("Actual 2024 Tax Revenue:", actual_2024, "\n")
cat("Percent Off:", round(percent_off_rf, 2), "%\n")

# Variable Importance
cat("\nVariable Importance:\n")
print(importance(rf_model))


```


##################### MODEL TWO - RANDOM FOREST - OPTIMIZATION ATTEMPT - PERFORMANCE IMPROVED  #####################



```{r}
library(caret)

# Define train control with 5-fold cross-validation
train_control <- trainControl(method = "cv", number = 5)

# Initialize variables for multiple runs
num_runs <- 10
results <- data.frame(MAE = numeric(num_runs), MSE = numeric(num_runs), Percent_Off = numeric(num_runs))
best_params <- list()

# Perform multiple runs
for (i in 1:num_runs) {
  set.seed(i)  # Ensure reproducibility for each run

  # Train the Random Forest model using caret with ranger
  caret_rf_model <- train(
    total_collected ~ gdp_value + unemployment + cpi,
    data = training_data_rf,
    method = "ranger",  # Faster Random Forest implementation
    trControl = train_control,
    tuneGrid = expand.grid(
      mtry = c(1, 2, 3), 
      splitrule = "variance", 
      min.node.size = c(1, 5, 10)
    )
  )

  # Predict for 2024
  predicted_2024_rf_optimized <- predict(caret_rf_model, newdata = test_data_2024)
  percent_off_rf_optimized <- abs(predicted_2024_rf_optimized - actual_2024) / actual_2024 * 100

  # Calculate training metrics
  caret_training_predictions <- predict(caret_rf_model, newdata = training_data_rf)
  mae_rf_optimized <- mean(abs(caret_training_predictions - training_data_rf$total_collected))
  mse_rf_optimized <- mean((caret_training_predictions - training_data_rf$total_collected)^2)

  # Store results and best parameters
  results[i, ] <- c(mae_rf_optimized, mse_rf_optimized, percent_off_rf_optimized)
  best_params[[i]] <- caret_rf_model$bestTune
}

# Summarize results across runs
mean_mae <- mean(results$MAE)
mean_mse <- mean(results$MSE)
mean_percent_off <- mean(results$Percent_Off)
sd_percent_off <- sd(results$Percent_Off)


cat("Caret Optimized Random Forest Regression\n")
cat("Training Metrics (Averages Across Runs):\n")
cat("Mean MAE:", round(mean_mae, 2), "\n")
cat("Mean MSE:", round(mean_mse, 2), "\n")
cat("\n2024 Prediction:\n")
cat("Mean Percent Off Across Runs:", round(mean_percent_off, 2), "%\n")
cat("Standard Deviation of Percent Off Across Runs:", round(sd_percent_off, 2), "%\n")

# Print Best Parameters Selected in Each Run
cat("\nBest Parameters Selected Across Runs:\n")
print(do.call(rbind, best_params))

```


```{r}
# Focus on the last 3 years and the 2024 prediction
focus_years <- c(2021, 2022, 2023, 2024)
focus_actual <- ny_tax_data$total_collected[ny_tax_data$fiscal_year %in% focus_years]

# Extract predictions for Random Forest models
focus_rf <- c(focus_actual[1:3], predicted_2024_rf)                     # Original Random Forest
focus_rf_optimized <- c(focus_actual[1:3], predicted_2024_rf_optimized) # Optimized Random Forest

windowsFonts(A = windowsFont("Times New Roman"))
par(family = "A")

plot(focus_years, focus_actual / 1e9, type = "l", col = "black", lwd = 2,
     ylim = c(min(focus_actual) * 0.95 / 1e9, max(focus_actual) * 1.05 / 1e9), 
     xlab = "Fiscal Year", ylab = "Tax Revenue (Billions $)", 
     main = "Random Forest: Actual vs. Original & Optimized", xaxt = "n", yaxt = "n")
grid()

lines(focus_years, focus_rf / 1e9, col = "red", lwd = 2, lty = 2)         # Original Model
lines(focus_years, focus_rf_optimized / 1e9, col = "blue", lwd = 2, lty = 3) # Optimized Model

axis(1, at = focus_years, labels = focus_years)
axis(2, at = seq(floor(min(focus_actual) / 1e9 * 0.95), 
                 ceiling(max(focus_actual) / 1e9 * 1.05), by = 5), 
     labels = paste0(seq(floor(min(focus_actual) / 1e9 * 0.95), 
                         ceiling(max(focus_actual) / 1e9 * 1.05), by = 5), "B"), 
     las = 1)

legend("topright", legend = c("Actual", "Original Model", "Optimized Model"), 
       col = c("black", "red", "blue"), lty = c(1, 2, 3), lwd = 2, cex = 0.7)

```







##################### MODEL THREE - EXPONENTIAL SMOOTHING MODEL OPTIMIZATION - ORIGINAL MODEL REINFORCED) #####################
```{r}
# Load necessary library
library(forecast)

# Define Time Series
tax_revenue_ts <- ts(ny_tax_data$total_collected[ny_tax_data$fiscal_year < 2024], 
                     start = 1995, frequency = 1)

# Fit ETS Model
ets_model <- ets(tax_revenue_ts)

# Training Metrics
training_predictions_ets <- fitted(ets_model)
mae_ets <- mean(abs(training_predictions_ets - tax_revenue_ts))
mse_ets <- mean((training_predictions_ets - tax_revenue_ts)^2)

# Forecast for 2024
forecast_2024_ets <- forecast(ets_model, h = 1)
predicted_2024_ets <- as.numeric(forecast_2024_ets$mean)
actual_2024 <- ny_tax_data$total_collected[ny_tax_data$fiscal_year == 2024]
percent_off_ets <- abs(predicted_2024_ets - actual_2024) / actual_2024 * 100

# Print Results
cat("Model Three: Regular ETS\n")
cat("Training Metrics:\n")
cat("MAE:", round(mae_ets, 2), "\n")
cat("MSE:", round(mse_ets, 2), "\n")
cat("\n2024 Prediction:\n")
cat("Predicted 2024 Tax Revenue:", round(predicted_2024_ets, 2), "\n")
cat("Actual 2024 Tax Revenue:", actual_2024, "\n")
cat("Percent Off:", round(percent_off_ets, 2), "%\n")

```

##################### MODEL 3 - EXPONENTIAL SMOOTHING MODEL-  #####################
```{r}
# Define all possible ETS configurations
ets_configs <- expand.grid(
  error = c("A", "M"),
  trend = c("N", "A", "M"),
  seasonal = c("N", "A", "M"),
  stringsAsFactors = FALSE
)

# Initialize variables to store best results
best_mae <- Inf
best_model <- NULL
results <- data.frame()

# Iterate through configurations
for (i in 1:nrow(ets_configs)) {
  config <- ets_configs[i, ]
  
  # Try to fit ETS model with current configuration
  tryCatch({
    model <- ets(
      tax_revenue_ts,
      model = paste(config$error, config$trend, config$seasonal, sep = "")
    )
    
    # Calculate training metrics
    training_predictions <- fitted(model)
    mae <- mean(abs(training_predictions - tax_revenue_ts))
    mse <- mean((training_predictions - tax_revenue_ts)^2)
    
    # Update the best model if MAE improves
    if (mae < best_mae) {
      best_mae <- mae
      best_model <- model
    }
    
    # Store results for each configuration
    results <- rbind(results, cbind(config, MAE = mae, MSE = mse))
  }, error = function(e) {
    # Handle models that fail to fit
    NULL
  })
}

# Final optimized ETS model
training_predictions_ets_optimized <- fitted(best_model)
mae_ets_optimized <- mean(abs(training_predictions_ets_optimized - tax_revenue_ts))
mse_ets_optimized <- mean((training_predictions_ets_optimized - tax_revenue_ts)^2)

# Forecast for 2024
forecast_2024_ets_optimized <- forecast(best_model, h = 1)
predicted_2024_ets_optimized <- as.numeric(forecast_2024_ets_optimized$mean)
percent_off_ets_optimized <- abs(predicted_2024_ets_optimized - actual_2024) / actual_2024 * 100


cat("\nModel Three: Optimized ETS\n")
cat("Training Metrics:\n")
cat("MAE:", round(mae_ets_optimized, 2), "\n")
cat("MSE:", round(mse_ets_optimized, 2), "\n")
cat("\n2024 Prediction:\n")
cat("Predicted 2024 Tax Revenue:", round(predicted_2024_ets_optimized, 2), "\n")
cat("Actual 2024 Tax Revenue:", actual_2024, "\n")
cat("Percent Off:", round(percent_off_ets_optimized, 2), "%\n")

# Print Best Configuration
cat("\nBest ETS Configuration:\n")
print(best_model)

```

```{r}

# Focus on the last 3 years and the 2024 prediction
focus_years <- c(2021, 2022, 2023, 2024)
focus_actual <- ny_tax_data$total_collected[ny_tax_data$fiscal_year %in% focus_years]

# Extract predictions for Original and Optimized ETS models
focus_ets <- c(focus_actual[1:3], predicted_2024_ets)
focus_ets_optimized <- c(focus_actual[1:3], predicted_2024_ets_optimized)


windowsFonts(A = windowsFont("Times New Roman"))
par(family = "A")

plot(focus_years, focus_actual / 1e9, type = "l", col = "black", lwd = 2,
     ylim = c(min(focus_actual) * 0.95 / 1e9, max(focus_actual) * 1.05 / 1e9), 
     xlab = "Fiscal Year", ylab = "Tax Revenue (Billions $)", 
     main = "ETS: Actual vs. Original & Optimized", xaxt = "n", yaxt = "n")
grid()


lines(focus_years, focus_ets / 1e9, col = "red", lwd = 2, lty = 2)           # Original Model
lines(focus_years, focus_ets_optimized / 1e9, col = "blue", lwd = 2, lty = 3)  # Optimized Model


axis(1, at = focus_years, labels = focus_years)
axis(2, at = seq(floor(min(focus_actual) / 1e9 * 0.95), 
                 ceiling(max(focus_actual) / 1e9 * 1.05), by = 5), 
     labels = paste0(seq(floor(min(focus_actual) / 1e9 * 0.95), 
                         ceiling(max(focus_actual) / 1e9 * 1.05), by = 5), "B"), 
     las = 1)


legend("topright", legend = c("Actual", "Original ETS", "Optimized ETS"), 
       col = c("black", "red", "blue"), lty = c(1, 2, 3), lwd = 2, cex = 0.7)


```


